\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-22 tex/paper.tex
save
losses are in expectation over X

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Section~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{Y|X} [ -\log p_{\theta_f} (Y|X) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{Z|f(X;\theta_f)} [-\log p_{\theta_r} (Z|f(X;\theta_f))].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

@@

\begin{proposition}\label{prop:1}
Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$ for fixed $\theta_f$. If $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
The optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and implies that $f$ is a pivotal quantity.
\end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = H(p_{Y|X}) - H(p_Z)$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the entropy
$H(p_{Z|f(X;\theta_f)})$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H(p_{Y|X}) - H(p_Z) \leq {\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $H(p_{Y|X})$,

    \item $\smash{\hat\theta}_f$ maximimizes the entropy
    $H(p_{Z|f(X;\theta_f)})$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely and in which case
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, which induces that the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity as a consequence of Proposition~\ref{prop:1}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - H(p_{Z|f(X;\theta_f)})$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $H(p_{Y|X}) - H(p_Z) < {\cal L}_f(\theta_f) -
H(p_{Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

>>>>

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Section~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{X} \mathbb{E}_{Y|X} [ -\log p_{\theta_f} (Y|X) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{X}  \mathbb{E}_{Z|f(X;\theta_f)} [-\log p_{\theta_r} (Z|f(X;\theta_f))].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

@@

\begin{proposition}\label{prop:1}
Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$ for fixed $\theta_f$. If $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
The optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ and implies that $f$ is a pivotal quantity.
\end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = \mathbb{E}_{X} [H(p_{Y|X}) - H(p_Z)]$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{X} [ H(p_{Z|f(X;\theta_f)}) ]$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - \mathbb{E}_{X} [ H(p_{Z|f(X;\theta_f)}) ].$$  In particular, we have the lower
bound $$\mathbb{E}_{X} [ H(p_{Y|X}) - H(p_Z)] \leq {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $\mathbb{E}_{X} [H(p_{Y|X})]$,

    \item $\smash{\hat\theta}_f$ maximimizes the expected entropy
    $\mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely. In other words,
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, which induces that the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity as a consequence of Proposition~\ref{prop:1}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - \mathbb{E}_X [ H(p_{Z|f(X;\theta_f)}) ]$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $\mathbb{E}_{X} [ H(p_{Y|X}) - H(p_Z)] < {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H(p_{Z|f(X;\theta_f)})]$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.
